Gene expression profiling of reproductive meristem identities in rice inflorescence

### 2015-12-15 ms.Rmd
Stefan and Helene's comments from lab meeting Added acknowledgements (fixes #32) and more biology to the conclusion (fixes #31) Shorter title (fixes #25)

---
title: Changes in gene expression associated with transitions in reproductive meristem identity during early development of rice inflorescence.
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage[utf8]{inputenc}
  \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  BiocStyle::html_document:
    dev: CairoPDF
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
    toc: yes
bibliography: bib/papersLib.bibtex
---

@@

@@

In rice, inflorescence architecture is established at early stages of reproductive development by the activity of different meristem types, and contributes directly to grain yield potential. After induction of flowering, the shoot apical meristem is converted to an inflorescence meristem to form the rachis, which is the main axis of the inflorescence. The rachis meristem produces primary branch meristems, which themselves produce axillary meristems during elongation. The axillary meristems may become secondary or higher-order branch meristems, retaining their indeterminate state, or be converted directly to determinate spikelet meristems. After elongation, the primary and higher-order branch meristems differentiate into spikelet meristems, and the complexity of branching and number of spikelets is fixed. Although some of the genes involved in transitions in meristem identity in rice have been identified, understanding of the network of genes controlling this process is lacking. To address this, we used a precise laser microdissection and RNA sequencing approach in *Oryza sativa* ssp. *japonica* cv. Nipponbare to produce quantitative data that describe the landscape of gene expression in four different meristem types. A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed and a number of potential candidates were identified. This resource will be vital for a mechanistic understanding of the link between inflorescence development and grain yield that integrates gene expression with environmental and other epigenetic factors, as well as for comparative studies of the domestication process.

@@

@@

The branched inflorescence of rice is a compound raceme, classified as a panicle, composed of a rachis (the main axis), primary branches, higher order branches, and spikelets. During panicle development, the undifferentiated cells in the primary axis (*i.e.* the rachis meristem, RM) undergo transitions through a series of meristem identities. After the establishment of the rachis some RM cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches. During elongation, the PBM is indeterminate and can give rise to axillary meristems (AM), which may differentiate into spikelets or secondary and higher-order branches. The PBM and secondary branch meristem (SBM) are both converted to a terminal spikelet meristem (SM). Terminal and axillary SMs develop into floret meristems (FM) and produce a single floret, meaning that grain yield is determined by the number of spikelets [@Ikeda:2004tk; @Yoshida:2011bo]. After spikelet differentiation, panicle branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk].

The landscape of gene expression is a characteristic factor that differentiates meristem cell types, and is involved in the control of transitions between meristem identities and therefore panicle phenotype. Mutant analyses and map-based cloning of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq]. Some of these genes are involved in the patterning of axillary meristem and panicle branching, such as the nuclear regulatory factor-encoding genes *MONOCULM 1*, *LAX PANICLE 1* and *LAX PANICLE 2* [@Wang:2011kq]. A large set of genes related to floral development, which may also affect panicle architecture, has been identified [@Yoshida:2011bo], some of which may have been under selection during domestication or are associated with crop improvement in Asian rice  [@Xing:2010de; @He:2011io; @Wang:2011kq; @Xu:2012hq; @Ikeda:2013go].

Understanding the events that occur during development that determine panicle characteristics, such as branching complexity and its plasticity, will be vital for sustainable improvement of rice yield potential using targeted breeding programs. Despite advances in characterization at the level of individual genes and interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and the environment [@Azpeitia:2013jia]. One step towards this goal is to describe the changes in gene expression between different cell types during development. In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *O. sativa* ssp. *japonica* cv. Nipponbare, using a precise laser microdissection and RNA sequencing approach.

@@

At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA. Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage. As a preliminary assessment of the RNA sequencing (RNAseq) results, the qualitative and quantitative expression of `r nIS` genes were compared with patterns that had previously been reported. `r nPos` of the `r nIS` genes were detected in two or more libraries from at least one meristem type in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`). The expression of further four genes that were detected in specific meristem types by RNAseq was confirmed by RNA *in situ* hybridization. Each of the four genes was detected in the same meristem types by both methods (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`).

@@

To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` software package [@Love:2014do]. Detected genes were ranked by variance, and standardized, VST-transformed read counts for the `r as.character(dim(expressionMatrix)[1])` most-variable genes (33% of expressed genes) were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw]. To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPcaCentroids"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`). Several common patterns were recovered by clustering: genes that increase or decrease in expression steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between the apical meristem (RM) and axillary meristem (PBM–SM) samples (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`) and genes that change expression gradually over the course of changes in axillary meristem identity (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`). There were also two weaker clusters containing fewer genes, which had complex expression patterns involving changes in expression particular to the PBM (clusters `r getClusterName('inc', 'dec')` and `r getClusterName('dec', 'inc')`).

@@

Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`). Unlike soft clustering, which was used to group genes based on shared expression dynamics, this analysis was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in read count of their members. Some  modes of expression shared between different families may suggest roles in inflorescence development. The most pronounced change in expression appears to occur between the RM and branch meristem samples (*i.e.* the change from apical to axillary meristem). Several families (including ABI3VP1 and GROWTH-REGULATING FACTOR TF genes and Aux/IAA genes) are more highly expressed in RM than in the other meristem types. Another group, including MYB and SBP TF genes and SET and PHD regulator genes, is enriched in one or both of the PBM and ePBM/AM samples but depleted or not enriched in the RM samples.

Homeodomain genes are involved at various stages of plant development and in several hormone response pathways (**ref Chan 1998**; **ref Himmelbach 2002**; **ref Sawa 2002**). The expression of this family in inflorescence meristems was explored in greater detail using a heatmap of scaled, transformed read counts. Heirarchical clustering of the transformed counts recovered five prominent groups of expression,  with unequal distribution of subfamilies of homeodomain genes among the groups (`r I(pasteLabel("Figure ", figCount, "f_hb"))`).

@@

Two results are presented here that suggest significant changes in gene expression between apical and axillary meristems: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM. Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). Axillary meristem initiation requires auxin synthesis and transport in *A. thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. In *A. thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], and transgenic rice plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co], have smaller panicles with fewer branches [@Yadav:2011db]. GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling. Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.

@@

Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM. The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity. *AP1* also interacts with members of the *AGL2*-like clade (*SEPALLATA* MADS-domain factors), *APETALA3* (*AP3*) and *PISTILLATA* (*PI*) to specify petal identity, and *AGL2*-like clade members also interact with *AGAMOUS* (*AG*) to specify stamens [recently reviewed by @Prunet:2013ki]. In rice, functions similar to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy], whilst stamen and carpel identity are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br; **ref Dreni 2011**]. Two recent studies in *A. thaliana* and the domesticated tomato, *Solanum lycopersicum*, used LMD and RNA sequencing to study inflorescence development [@Park:2012dg; @Mantegazza:2014di]. Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^ genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`). *A. thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems. *agl6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de]. The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems. Expression of the putative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage. In contrast, consistent with their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.

The strong enrichment of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the PBM and their presence in cluster `r getClusterName('small', 'dec')` suggests involvement in the early stages of inflorescence development. *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c), suggesting a redundant function in indeterminate axillary meristems. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx], *SPL* gene expression is controlled by *miR156* and *miR529* [**ref Jeong 2011**; @Wang:2015jx]. The precursor for another microRNA, *MIR319A*, was recovered from the same cluster as *SPL3* and *SPL12* and has a similar expression pattern to these genes, *SPL17* and *WFP* (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c). Overexpression of *WFP*, which was detected by *in situ* hybridisation experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [@Lu:2013hy], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O. sativa* ssp. *japonica* varieties. Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of *GN1A* [@Huang:2009kf]. WFP also binds to the promoters of *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].

>>>>

---
title: Gene expression profiling of reproductive meristem identities in rice inflorescence
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage[utf8]{inputenc}
  \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
  BiocStyle::html_document:
    dev: CairoPDF
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
bibliography: bib/papersLib.bibtex
---

@@

Expression profile of rice inflorescence meristems

@@

In rice, inflorescence architecture is established at early stages of reproductive development by the activity of different meristem types, and contributes directly to grain yield potential. After induction of flowering, the shoot apical meristem is converted to an inflorescence meristem to form the rachis, which is the main axis of the inflorescence. The rachis meristem produces primary branch meristems, which themselves produce further axillary meristems during elongation. The axillary meristems may become secondary or higher-order branch meristems, retaining their indeterminate state, or be converted directly to determinate spikelet meristems. The timing of rachis meristem abortion specifies the number of primary branches, whilst the transition of indeterminate branch meristems to determinate spikelet meristems conditions the complexity of branching and thereby the number of seeds on the panicle. Although some of the genes involved in transitions in meristem identity in rice have been identified, understanding of the network of genes controlling this process is lacking. To address this, we used a precise laser microdissection and RNA sequencing approach in *Oryza sativa* ssp. *japonica* cv. Nipponbare to produce quantitative data that describe the landscape of gene expression in four different meristem types. A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed and a number of potential candidates were identified. This resource will be vital for a mechanistic understanding of the link between inflorescence development and grain yield that integrates gene expression with environmental and other epigenetic factors, as well as for comparative studies of the domestication process.

@@

Rice inflorescence meristems undergo a series of transitions during development, and control of these transitions impacts the morphology of the adult inflorescence and hence grain yield. We use laser microdissection and RNA sequencing to describe the transcriptional changes that occur during this process.

@@

The branched inflorescence of rice is a compound raceme, classified as a panicle, composed of a rachis (the main axis), primary branches, higher order branches, and spikelets. During panicle development, the undifferentiated cells in the primary axis (*i.e.* the rachis meristem, RM) undergo transitions through a series of meristem identities. After the establishment of the rachis some RM cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches. During primary branch elongation, the PBM is indeterminate and can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelets. The PBM and secondary branch meristem (SBM) are both converted to a terminal spikelet meristem (SM). Terminal and axillary SMs develop into floret meristems (FM) and produce a single floret, meaning that grain yield is determined by the number of spikelets [@Ikeda:2004tk; @Yoshida:2011bo]. After spikelet differentiation, panicle branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk].

The landscape of gene expression is a characteristic that differentiates meristem types, and is involved in the control of transitions between meristem identities and therefore panicle branching. Mutant analyses and map-based cloning of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq]. Some of these genes are involved in the patterning of axillary meristem and panicle branching, such as the nuclear regulatory factor-encoding genes *MONOCULM 1*, *LAX PANICLE 1* and *LAX PANICLE 2* [@Wang:2011kq]. A large set of genes related to floral development, which may also affect panicle architecture, has been identified [@Yoshida:2011bo], some of which may have been under selection during domestication or are associated with crop improvement in Asian rice  [@Xing:2010de; @He:2011io; @Wang:2011kq; @Xu:2012hq; @Ikeda:2013go].

Understanding the events that occur during development that determine panicle characteristics, such as branching complexity and its plasticity, will be vital for sustainable improvement of rice yield potential using targeted breeding programs. Despite advances in characterization at the level of individual genes and interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and the environment [@Azpeitia:2013jia]. One step towards this goal is to describe the changes in gene expression between different meristem types during development. In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *O. sativa* ssp. *japonica* cv. Nipponbare, using a precise laser microdissection and RNA sequencing approach.

@@

At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA. Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage. As a preliminary assessment of the RNA sequencing (RNAseq) results, the qualitative and quantitative expression of `r nIS` genes were compared with patterns that had previously been reported. `r nPos` of the `r nIS` genes were detected in two or more libraries from at least one meristem type in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`). The expression of a further four uncharacterized genes that were detected in specific meristem types by RNAseq was confirmed by RNA *in situ* hybridization. Each of the four genes was detected in the same meristem types by both methods (`r I(pasteLabel("Figure S", s_figCount, "sf_inSitu"))`). These results suggest that the tissues used for RNA sequencing were accurately dissected and represent the intended meristem types.

@@

To recover common expression patterns in expressed genes, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` software package [@Love:2014do]. Detected genes were ranked by variance, and standardized, VST-transformed read counts for the `r as.character(dim(expressionMatrix)[1])` most-variable genes (33% of expressed genes) were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Kumar:2007uw]. To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figure S", s_figCount, "sf_mfuzzPcaCentroids"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`). Several common patterns were recovered by clustering: genes that increase or decrease in expression steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between the apical meristem (RM) and axillary meristem (PBM to SM) samples (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`) and genes that change expression gradually over the course of changes in axillary meristem identity (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`). There were also two weaker clusters containing fewer genes, which had complex expression patterns involving changes in expression particular to the PBM (clusters `r getClusterName('inc', 'dec')` and `r getClusterName('dec', 'inc')`).

@@

Because of the enrichment of TF family members in some of the gene expression clusters, geneset enrichment analysis (GSEA) was used to provide an overview of expression of transcription factors and other regulators in the meristem samples (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`). Unlike soft clustering, which was used to group genes based on shared expression dynamics, this analysis was used to investigate the overall strength of expression of the families based on Log~2~-fold changes (L~2~FC) in read count of their members. The most pronounced change in expression appears to occur between the RM and branch meristem samples (*i.e.* the change from apical to axillary meristem). Several families (including ABI3VP1 and GROWTH-REGULATING FACTOR TF genes and Aux/IAA genes) are more highly expressed in RM than in the other meristem types. Another group, including MYB and SBP TF genes and SET and PHD regulator genes, is enriched in one or both of the PBM and ePBM/AM samples but depleted or not enriched in the RM samples.

Homeodomain genes are involved at various stages of plant development and in several hormone response pathways (**ref Chan 1998**; **ref Himmelbach 2002**; **ref Sawa 2002**). The expression of this family in inflorescence meristem types was explored in greater detail using a heatmap of scaled, transformed read counts. Hierarchical clustering of the transformed counts recovered five prominent groups of expression (`r I(pasteLabel("Figure ", figCount, "f_hb"))`). There was no clear relationship between homeodomain subfamily and expression pattern except for an apparent enrichment of class IV HD-Zip genes with a pattern of lowest expression in the RM, a peak in indeterminate axillary meristems and then a gradual decrease during axillary meristem transitions.

@@

Significant changes in gene expression between apical and axillary meristems are supported by two observations: a difference in expression of annotated transcription factor genes between RM and PBM samples, and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM. Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM–SM). Axillary meristem initiation requires auxin synthesis and transport in *A. thaliana* and maize [reviewed by @Gallavotti:2013ki], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters, while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. In *A. thaliana*, excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], and transgenic rice plants containing 3 kb of upstream sequence from *FLO-LFY HOMOLOG OF RICE* (*RFL*) fused to the *GH3-8* coding sequence, presumably expressing *GH3-8* in the endogenous pattern of *RFL* [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co], have smaller panicles with fewer branches [@Yadav:2011db]. GH3-8 enzyme conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling. Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type, and understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues.

@@

Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM, which corresponds to the acquisition of determinate fate by the axillary meristem. The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` corresponds with their involvement in flower development in rice and other plants. In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity. *AP1* also interacts with members of the *AGL2*-like clade (*SEPALLATA* MADS-domain factors), *APETALA3* (*AP3*) and *PISTILLATA* (*PI*) to specify petal identity, and *AGL2*-like clade members also interact with *AGAMOUS* (*AG*) to specify stamens [recently reviewed by @Prunet:2013ki]. In rice, functions similar to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy], whilst stamen and carpel identity are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br; **ref Dreni 2011**]. Two recent studies in *A. thaliana* and the domesticated tomato, *Solanum lycopersicum*, used an LMD and RNAseq approach to investigate inflorescence development [@Park:2012dg; @Mantegazza:2014di]. Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^ genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`). *A. thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems. *agl6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de]. The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems. Expression of the putative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage. In contrast, consistent with their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.

The strong enrichment of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the PBM and their presence in cluster `r getClusterName('small', 'dec')` suggests involvement in the early stages of inflorescence development. *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c), suggesting a redundant function in indeterminate axillary meristems. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx], *SPL* gene expression is controlled by *miR156* and *miR529* [**ref Jeong 2011**; @Wang:2015jx]. The precursor for another microRNA, *MIR319A*, was recovered from the same cluster as *SPL3* and *SPL12* and has a similar expression pattern to these genes, *SPL17* and *WFP* (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c). In *A. thaliana*, *miR319a* targets *TEOSINTE BRANCHED/CYCLOIDEA/PROLIFERATING CELL FACTORS* (*TCP*) transcription factors and is involved in floral organ development (**ref Palatnik 2003**; **ref Nag 2009**), but no direct interaction with *SPL* genes or role in reproductive development has been reported for *mrR319a* in rice. Overexpression of *WFP*, which was detected by *in situ* hybridization experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [@Lu:2013hy], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O. sativa* ssp. *japonica* varieties. Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*) [@Huang:2009kf]. WFP also binds to the promoters of *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].

A homolog of *LOG1*, *LONELY GUY LIKE PHOSPHORIBOHYDROLASE 1* (*LOGL1*), is present in cluster `r getClusterName('inc', 'inc')` with genes that increase in expression between RM and SM. In rice, cytokinins may increase branching complexity by promoting IM or BM activity [reviewed by @Han:2014dt]. The LOG1 enzyme converts cytokinin nucleotides to the active form *in vitro*, and *log1* mutants have lower expression of two cytokinin-inducible *RESPONSE REGULATOR* genes, resulting in early termination of BM and IM and small panicles with small SAMs and branching defects [@Kurakawa:2007go]. Similarly, reduced expression *GN1A*, which encodes an enzyme that inactivates cytokinin, is associated with increased grain production in several high-yield *O. sativa* ssp. *indica* varieties [@Ashikari:2005eg]. Although *GN1A* transcripts were not detected in this LMD dataset (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`d), these results suggest that an increase in active cytokinin is associated with grain yield via an effect on meristem activity, but the *LONG AND BARBED AWN1* (*LABA1*/*LOGL6*/*An-2*) gene provides evidence for an inverse role. The LABA1 protein present in *O. rufipogon*, which shares an identical primary sequence with the protein from *O. sativa* ssp. *japonica* cv. Nipponbare, also activates cytokinin *in vitro* [@Gu:2015hy; @Hua:2015bs], and introduction of the *Laba1* allele from *O. rufipogon* into an *O. sativa* ssp *indica* background that contains a non-functional *laba1* allele results in a higher concentration of endogenous cytokinins and higher expression of *RESPONSE REGULATOR* genes [@Hua:2015bs], but reduces the number of grains per panicle and the number of tillers per plant [@Gu:2015hy]. *laba1* is a domestication allele found at high frequencies in cultivated accessions, implying that it was affected by artificial selection either for awn phenotype or grain yield [@Gu:2015hy; @Hua:2015bs]. *LABA1* was detected in the ePBM/AM and SM samples but not the RM or PBM samples. Although the function of LOGL1 has not been reported, the expression patterns of *LABA1* and *LOGL1* suggest increasing cytokinin activation along the course of axillary meristem determination (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`d). Such a mechanism would support a cytokinin function secondary to the maintenance of IM and BM, but it is currently not clear how this would influence meristem identity or panicle branching at the molecular level. A type-A *RESPONSE REGULATOR* (*RR3*) is present in cluster `r getClusterName('dec', 'small')` (containing genes more highly expressed in RM), highlighting the possible complexity of the roles of cytokinins in inflorescence development.
